% Deep Emergence Experiment Configuration for Manuscript
% Copy this into your LaTeX document

\subsection{Experimental Configuration}

We conducted large-scale meta-agent emergence experiments using the configuration detailed in Table~\ref{tab:deep_emergence_config}. The experimental setup was designed to study deep hierarchical emergence across many scales with aggressive consensus dynamics and multi-level hyperprior propagation (Ouroboros Tower).

\begin{table}[htbp]
\centering
\caption{Deep Emergence Experiment Configuration}
\label{tab:deep_emergence_config}
\begin{tabular}{llr}
\toprule
\textbf{Category} & \textbf{Parameter} & \textbf{Value} \\
\midrule
\multicolumn{3}{l}{\textit{System Initialization}} \\
& Initial agents ($N_0$) & 8 \\
& Latent dimension ($K$) & 13 \\
& Observation dimension ($D_x$) & 5 \\
& Random seed & 2 \\
\midrule
\multicolumn{3}{l}{\textit{Evolution Dynamics}} \\
& Maximum steps & 500 \\
& Logging interval & 1 \\
& Stop if scales reached & 25 \\
& Stop if max active agents & 200 \\
\midrule
\multicolumn{3}{l}{\textit{Emergence Mechanism}} \\
& Consensus threshold ($\tau_{\text{KL}}$) & 0.05 \\
& Consensus check interval & 2 steps \\
& Minimum cluster size & 2 \\
& Maximum scale ($\zeta_{\max}$) & 20 \\
& Max constituents per meta-agent & 10 \\
& Max total agents (hard cap) & 1000 \\
\midrule
\multicolumn{3}{l}{\textit{Hierarchical Architecture}} \\
& Cross-scale priors & Enabled \\
& Ouroboros Tower & Enabled \\
& Hyperprior depth & 5 levels \\
& Hyperprior decay ($\gamma$) & 0.5 \\
\midrule
\multicolumn{3}{l}{\textit{Energy Weights}} \\
& Self-energy ($\lambda_{\text{self}}$) & 1.0 \\
& Belief alignment ($\lambda_{\text{belief}}$) & 1.0 \\
& Prior alignment ($\lambda_{\text{prior}}$) & 1.0 \\
& Observation ($\lambda_{\text{obs}}$) & 1.0 \\
& Gauge coupling ($\lambda_{\phi}$) & 1.0 \\
& Softmax temperature ($\kappa_\beta, \kappa_\gamma$) & 1.0 \\
\midrule
\multicolumn{3}{l}{\textit{Learning Rates}} \\
& Belief mean ($\eta_{\mu_q}$) & 0.05 \\
& Belief covariance ($\eta_{\Sigma_q}$) & 0.0075 \\
& Prior mean ($\eta_{\mu_p}$) & 0.02 \\
& Prior covariance ($\eta_{\Sigma_p}$) & 0.0075 \\
& Gauge field ($\eta_{\phi}$) & 0.1 \\
\midrule
\multicolumn{3}{l}{\textit{Diagnostics}} \\
& Meta-agent visualizations & Enabled \\
& Snapshot interval & 1 step \\
& Diagnostic plots & Enabled \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Design Rationale}

The configuration implements several key design choices:

\paragraph{Aggressive Emergence.}
Consensus checks occur every 2 steps (compared to 5 in standard configurations), enabling rapid hierarchical condensation. The low consensus threshold ($\tau_{\text{KL}} = 0.05$) allows meta-agent formation when agents achieve sufficient epistemic alignment.

\paragraph{Ouroboros Tower Architecture.}
The 5-level hyperprior propagation implements Wheeler's ``it from bit'' concept extended to hierarchical systems. Each agent receives priors not only from its immediate parent meta-agent, but from ancestors up to 5 scales higher, with exponential decay $\gamma = 0.5$ to weight more recent ancestry.

\paragraph{Balanced Energy Landscape.}
All energy weights are set to unity ($\lambda_{\text{self}} = \lambda_{\text{belief}} = \lambda_{\text{prior}} = \lambda_{\text{obs}} = 1.0$), creating an unbiased energy landscape. This allows the system to explore the natural balance between individual identity preservation (self-energy), lateral social coordination (belief alignment), vertical hierarchical integration (prior alignment), and empirical grounding (observation fit). The absence of hierarchical bias enables observation of emergent dynamics without imposed preferences.

\paragraph{Safety Constraints.}
Early stopping conditions prevent computational runaway: evolution terminates upon reaching either 25 scales or 200 total active agents. These limits ensure tractable computation while allowing sufficient depth to observe multi-scale emergence dynamics.

\paragraph{Dense Temporal Sampling.}
Snapshots are captured at every step, providing high-resolution data for analyzing emergence events, information flow, and hierarchical condensation dynamics.

% Alternative compact version for space-constrained journals:

\begin{table}[htbp]
\centering
\caption{Deep Emergence Experiment Configuration (Compact)}
\label{tab:deep_emergence_compact}
\small
\begin{tabular}{lr|lr}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Parameter} & \textbf{Value} \\
\midrule
$N_0$ (initial agents) & 8 & $\lambda_{\text{self}}$ & 1.0 \\
$K$ (latent dim) & 13 & $\lambda_{\text{belief}}$ & 1.0 \\
$D_x$ (obs dim) & 5 & $\lambda_{\text{prior}}$ & 1.0 \\
Max steps & 500 & $\eta_{\mu_q}$ & 0.05 \\
$\tau_{\text{KL}}$ threshold & 0.05 & $\eta_{\Sigma_q}$ & 0.0075 \\
Consensus check & 2 steps & $\eta_{\mu_p}$ & 0.02 \\
$\zeta_{\max}$ (max scale) & 20 & $\eta_{\Sigma_p}$ & 0.0075 \\
Hyperprior depth & 5 & Snapshot interval & 1 \\
Hyperprior decay $\gamma$ & 0.5 & Stop at scales & 25 \\
\bottomrule
\end{tabular}
\end{table}

% For the methods section, you might also want this prose description:

\subsection{Simulation Procedure}

Simulations were initialized with $N_0 = 8$ agents in a $K=13$ dimensional latent space, each maintaining beliefs $q_i(c)$ and priors $p_i(c)$ as Gaussian distributions over the concept space $\mathcal{C}$. At each time step, agents updated their beliefs via natural gradient descent on the total free energy:
%
\begin{equation}
F_{\text{total}} = \sum_i \left[ \lambda_{\text{self}} F_{\text{self}}^{(i)} + \lambda_{\text{belief}} F_{\text{belief}}^{(i)} + \lambda_{\text{prior}} F_{\text{prior}}^{(i)} + \lambda_{\text{obs}} F_{\text{obs}}^{(i)} \right]
\end{equation}
%
where $F_{\text{self}}^{(i)} = \text{KL}(q_i \| p_i)$ measures intrinsic energy, $F_{\text{belief}}^{(i)}$ measures lateral alignment with neighbors, $F_{\text{prior}}^{(i)}$ measures vertical alignment with parent meta-agents, and $F_{\text{obs}}^{(i)}$ measures fit to external observations. With all energy weights set to unity ($\lambda_{\text{self}} = \lambda_{\text{belief}} = \lambda_{\text{prior}} = \lambda_{\text{obs}} = 1.0$), the system balances individual identity, social coordination, hierarchical integration, and empirical grounding without imposed bias.

Every 2 steps, the system checked for epistemic consensus by computing pairwise KL divergences $\text{KL}(q_i \| q_j)$ between all agents at each scale. When a cluster of agents satisfied $\text{KL}(q_i \| q_j) < \tau_{\text{KL}} = 0.05$, they underwent \textit{epistemic death} and condensed into a meta-agent at the next scale, which inherited a Gaussian prior synthesized from the cluster's belief distribution.

The Ouroboros Tower mechanism propagated priors down from ancestors up to 5 scales higher, weighted by exponential decay $\gamma^{\Delta\zeta}$ where $\Delta\zeta$ is the scale distance. This implements multi-level self-reference, allowing agents to be influenced by the emergent structure of the hierarchy itself.

Evolution continued until either (1) 25 scales emerged, (2) 200 total agents existed, or (3) 500 steps elapsed. All runs used random seed 2 for reproducibility. Comprehensive diagnostics were captured at every step, including energy decomposition, prior evolution, and hierarchical structure.
